---
title: |
  Double jeopardy
subtitle: The implications of gender on white's racial attitudes
short-title: Double jeopardy
published: "Working paper: please do not share without author permission*"
code-repo: "Replication materials are stored at <https://github.com/DamonCharlesRoberts/wid-and-gender>"
author:
  - name: Komal Preet Kaur
    email: komal.kaur@colorado.edu
    orcid: 0000-0001-9275-0906
    title: PhD Candidate
    affiliations: 
      - id: CU
        name: University of Colorado Boulder
        department: Political Science
        address: 333 UCB
        city: Boulder
        region: CO 
        postal-code: 80309-0333
    attributes:
      corresponding: false
  - name: Damon C. Roberts
    email: damon.roberts-1@colorado.edu
    orcid: 0000-0002-4360-3675
    title: PhD Candidate
    affiliations:
      - id: CU
        name: University of Colorado Boulder
        department: Political Science
        address: 333 UCB
        city: Boulder
        region: CO 
        postal-code: 80309-0333
    attributes:
      corresponding: true   
abstract: |
  Do White women report a higher racial identity than men? White political identity and co-racial attitudes recieve considerable scholarly and public attention in the post-Trump era. The field does not yet know why White men tend to hold more racially resentful views while White women tend to express more in-group oriented racial views. We argue that this gender gap reflects differences in what appeals are most attractive and that the palatability of these messages are governed by ideological differences. Using data from the 2016 and 2020 American National Election Studies (ANES), our evidence suggests that once we hold partisanship constant, Women report a higher White political identity than men. Using data from an experiment, we then examine the mechanism we argue underlies this phenomenon. Our results highlight the importance of considering gender-race identities in evaluating White identity in the United States. 
additional-info: |
  We have no known conflict of interest to disclose.
keywords:
  - white identity
  - racial attitudes
  - gender
  - elite messaging
date: today
bibliography: references.bib
format:
  hikmah-pdf: default
  
execute:
  echo: false
  warning: false
  message: false
---
{{< pagebreak >}}

# Introduction {#sec-introduction}

The in-group attitudes of whites garner significant attention by political pundits and by scholars. Many of these accounts document the tendency for the Republican party to represent those interests through positioning themselves on white victimhood [@samuels_lewis_2022_fte]. This appears to be a shift, however, from the observation that Republicans express racial animus. For example, during Obama's presidency, the Republican party used the increased salience of race relations to racialize any policy that had a connection to Obama, such as the Affordable Care Act [@tesler_2016]. So is it that Whites have shifted in what messages appeal to them: from messages of racial animus to white victimhood?

We argue that an important dimension of this that has not yet been accounted for in the literature is gender. In this manuscript, we argue that there is a gender gap in the palatability of racialized messages. As white women tend to be more liberal, when holding partisanship constant, we argue that messages from the party that express in-group preferences play better than racially resentful messages which overwhelmingly conjure images of a racist white man as the target audience. As white women are not necessarily immune from holding racialized views, we suspect that these messages do not yet have associations with staunchly conservative views nor have they been widely marked as racist and that this appeals to those who may self-identify as more liberal. We suspect that this provides a partial explanation as to how Donald J. Trump, a candidate who embodies a campaign of white victimhood, was so successful at capturing the vote among white women. 

Using data from the 2016 and 2020 American National Election Studies (ANES), we first establish whether there is a gender gap in white identification once we hold partisanship constant. From our analyses, we find that in both 2016 and 2020, that once we hold partisanship constant, that white women tend to report a higher white identity than white men. Next, in this draft, we provide a proposed experimental design that we believe will allow us to test the mechanism for this gender gap among whites. In the following section we elaborate on the mechanism that we argue explains this phenomenon.


# White identity versus racial animus

The conceptualization of white identity heavily relies on social identity theory. As scholars see it, white identity reflects a recognition of your connection to other whites and a preference for protecting the status of other whites [@helms_carter_1990;@jardina_2019_cup] - which implies a desire to protect the racial status quo [@schildkraut_2017_pgi]. 

Though race is heavily intertwined with partisan identification [@engelhardt_2020;@westwood_peterson_2022_pb], which is argued to be a central identity for Americans [@mason_2015_ajps], evidence suggests that both white Democrats and Republicans identify with their race; though white identity is a strong predictor of support for Trump in the 2016 election [@jardina_et-al_2020].

Whites protecting their status is not a new idea [see @painter_2010_ww], however, scholars are reconsidering the way that it manifests. The tendency for those studying the attitudes of white Americans is to examine their out-group attitudes. After all, racism in America has deep cultural and institutional roots [@lopez_2006_nyu;@painter_2010_ww;@rile-hayward_2013_cup;@al-ramiah_hewstone_2013]. Scholars in this area understand that racial animus, prejudice held toward non-whites [@kam_burge_2019], guides news coverage of American politics [@winter_2008], guides attitudes toward government spending [@gilens_1999; @hopkins_2009;@tesler_2016], guides attitudes toward vote choice [@sides_et-al_2018]; and guides the campaign strategies of candidates [@stephens-dougan_2020]. 

Racialized attitudes for whites depend on a number of factors. For example, from a social psychological perspective, both in-and-out-group racial attitudes are guided by a sense of status loss [@morrison_et-al_2009]. Whether this status loss is economic or is social and cultural is up for debate [see @roberts_2022_wp], many examples abound of whites' attitudes shifting when faced with a sense of change to the racial status quo [@bobo_1983;@albertson_gadarian_2015;@craig_richeson_2014]. Looking ascriptively, racially inclusive messages appear to be more popular among white woman relative to white men [@hutchings_et-al_2004_poq]. White men also were much less willing to support Obama relative to white women [@junn_2020_pop]. Some may assume that age matters, however, evidence suggests that once we adjust our measures of racialized attitudes, generational gaps disappear [@desante_smith_2020_chicago] and that the "old-fashioned racism" of the Jim Crow era remains [@tesler_2013].

This suggests that there are differences between men and women in whether in-group or out-group racial attitudes are more appealing. What is unclear, however, is what the mechanism that drives these differences is. In the next section, we make a case for one possible mechanism.

# The racialized gender gap

To understand why this racialized gender gap may exist, we need to consider how white men and women evaluate politics differently. A rich literature about gender documents a "gender gap"; the tendency for men and women to report meaningfully different political views. Women tend to be more liberal and Democratic, whereas men tend to be more conservative and Republican. A leading argument is that Republican men have taken a much more conservative stance on various issues and are responsible for widening this gender gap [@kaufmann_petrocik_1999_ajps]. Even looking within the parties, men tend to be much more conservative than women [@barnes_cassese_2017].

Many suggest that this gender gap exists as the result of the deep roots of the patriarchy in America. As white women experience marginalization, they rely on different considerations than white men. Specifically, white men relied more on their partisanship and anti-black attitudes to evaluate candidates in the 2018 midterm elections, whereas women relied more on their ideology and their sexism [@lemi_2022_jwpp]. One plausible explanation for this is a result of the historical differences in how white men and women are socialized. As America has deep roots in white supremacist ideology, white women are seen as symbols of purity and that they have a higher obligation to protecting the purity of the race [@junn_2017_pgi]. They face significant hostility by white men if they place their independence as women above their obligations to white men [@sunderland_2022_mm].

Though white women are marginalized, their experiences of marginalization do not translate into advocacy for marginalized racial groups. For example, white women hold negative views of the Black Lives Matter movement [@isom_et-al_2022_cs]. When advocacy groups designed for helping marginalized non-whites are lead by white women tend to be less active than if they are lead by non-white women [@strolovitch_2006_jop]. Though they are marginalized, this occurs because they still balance the considerations of their whiteness [@lucas_mohamed_2021_apr]. 

As white women tend to hold more liberal attitudes, rely on their ideology more than their partisanship, and experience marginalization themselves, we should expect that this racialized gap is the result of a balancing act between these considerations. White men are racially conservative and are amenable to accepting messages from Republicans that historically rely on messages of racial animus and stereotypes. White men do not experience marginalization, and as a result, changes to the racial status quo do not represent an increasing degree of victimization.

We expect the following: that once we control for partisanship, white women will report a higher white political identity than white men (**$H_1$**). The reason this racialized gender gap exists is that messages expressing the victimhood of whites engender stronger reactions among white women than men (**$H_2$**) and that this reaction to changes in the status quo are moderated by ideological views (**$H_3$**).

In the next section we first establish the existience of the racialized gender gap (**$H_1$**) using data from the 2016 and 2020 American National Election Studies. In this draft, we then propose a set of experimental studies that we intend to conduct to examine the plausibility of **$H_2$** and **$H_3$**. 

<!--

NOTE FOR KK: 
  - I LIKE ALL OF THIS! YOU'RE GOOD! I JUST DON'T FULLY AGREE WITH THE FRAME THAT THIS IS NEW. IT ISN'T, OUR DOCUMENTATION OF IT IS JUST NEW. SO I WANTED TO ADJUST THE FRAMING OF THE PAPER A LITTLE BIT. 
  - I ALSO THINK I FOUND A WAY TO KIND OF MERGE OUR TWO THEORIES TOGETHER!!! :)


White's heightened group consciousness is a relatively recent phenomenon. As earlier studies show, whites political attitudes and policy opinions were not always shaped by their racial identity [@sears2006political]. The Presidential election of Barack Obama followed by Donald Trump has made racial identity politically salient in American politics. Eight years of the nation's first Black president were followed by Donald Trump's explicitly racial rhetoric. His slogan of 'Make America great again' was about appealing to White voters to `restore' their dominance in politics. As Blumer posits in his theory of group position, these political events along with the movements such as Black Lives Matter, and demographic changes in the country have provided whites the information about their dwindling group status that has activated racial consciousness among Whites [@blumer1958race, @isom2022white, @mutz2018status]. 

The surge in scholarly work on White political identity has greatly improved our understanding of its implications on political behavior and policy preferences. Much of the existing work attempts to control for gender during empirical analysis and omits to delve deeper into the role of gender in political behavior. Therefore, we have limited knowledge of whether White political identity is gendered. This knowledge gap has theoretical and empirical implications, after all, social constructs such as race and gender are interdependent. Therefore, it is theoretically limiting to expect that racial identities work the same for all without accounting for gender differences. 

In this manuscript, we employ a gendered lens to understand ways in which the association with White identity differs for men and women. White men and women have different considerations for forming an association with their racial identity. We propose that White women tend to be more associated with their racial identity than White men. Drawing on social identity theories, particularly on race and gender, we explore the role of whether white women are more drawn to racial in-group messaging than White men.  
 

Using observational data from American National Election Studies in 2016 and 2020, we find that White women have a higher level of association with their racial identity than men. We propose an experimental approach for further testing underlying mechanisms.  

We make several contributions to the literature on women and politics as well as race and politics. One, we analyze one of the central questions on White identity and expand our understanding of racial attitudes of Whites particularly when the United States is becoming more racially diverse. Second, we attempt to parse out gendered and ideological dimensions of how racial identification translates into political preferences. As [@junn2017trump] argues that analyzing individual identity-level traits can offer fresh explanations for political decision-making, our manuscript is a step toward this direction. Moreover, white women are considered ``swing voters' [@junn2017trump] and are a relevant group that can make or break an election. Against this backdrop, our manuscript examines factors associated with White identity for women.  

In the following sections, we first review existing literature on white political identity and gendered political behavior that form our theoretical expectations. We then present our methodological approach to test our hypotheses. In the next section, we present our results from the observational data and a proposed experimental design. 

# Literature review

## Political and Policy Implications of White Identity

A growing number of whites have begun to feel threatened by the increasingly racially and ethnic diversity in the USA. As a result, they have a heightened sense of their racial group identification. Their sense of being discriminated against and the linked fate with other Whites are linked with whites' support for their descriptive representatives [@schildkraut2017white]. White American view movements such as Black Lives Matter as threatening their racial status quo, and feel victimized, which explains the support for Trump as a presidential candidate [@isom2022white]. Even though Whites Americans hold a dominant position socially, politically, and economically, they perceive that they are as much discriminated against as other groups in the country [@berry2022relative]. These perceptions further shape their electoral behavior [@berry2022relative]. 2016 Presidential elections    

Though conservatives are more likely to hold White identity dear to them than liberals [@jardina2019white], White identity is heightened in interracial contests [@petrow2018white]. Whites report linked fate more than other races that are associated with higher levels of political participation [@berry2019white]. Democrats associate with White identity when they feel insecure about the increasing racial diversity of the party [@perez_et-al_2021_pp]. White identity is also associated with the support of welfare policies when their in-group is the main benefactor [@jardina2019white].

White identity is not only politically relevant but also shapes policy preferences. For instance, @jardina2019white shows that White identity is associated with fewer preferences for immigration policies as a racially diverse environment threatens White dominance. Further, @perez_et-al_2021_pp demonstrate that White Democrats, despite being liberal, tend to oppose racially impactful policies such as immigration on economic grounds as they find cultural threats from immigration less palatable.  

The extant literature on identity shows that in-group and out-group attitudes have distinct political implications. When primed about in-group messaging, Whites tend to hold on to white identity more and therefore prefer a White than a Black presidential candidate [@petrow2018white]. Racial resentment, an out-group process,.............

## Linking Gender and White Political Identity

The examination of White women's political behavior has revealed somewhat puzzling patterns. White women tend to be more liberal than White men [CITE]. Earlier studies found that White women evaluate political candidates differently than men when they are exposed to racial cues. White women positively view Republicans' racially inclusive messages more than White men [@hutchings2004compassion]. However, unlike white men, Trump's support among white women was not rooted in the white victimhood narrative [@isom2022white]. But, their feeling of victimhood was manifested alongside their anti-Black Lives Matter stance.   

Though White democratic women overwhelmingly supported Obama, White women have been consistent supporters of Republican Presidential candidates more than men [@junn2020gender]. As opposed to women of color, hostile sexism -- perceiving women negatively who defy traditional gender norms -- positively predicts White women's preference for Trump [@frasure2018choosing].   


When it comes to intergroup relations, White women are less likely to support the political representation of non-White women [@matos2021trends]. When primed about gender discrimination, White women express pro-in-group bias and anti-out-group bias feelings [@craig2012unto]. Moreover, womanhood is important to White women and they tend to preserve White privilege [@junn2017trump].

Using data from the 2016 and 2020 American National Election Studies (ANES) and original data from a series of experiments, we test the claim that making their messages more palatable to relatively more liberal views helped the Republican party in attracting White women who may hold racially conservative views but were turned off by messages of racial resentment. The results demonstrate that ... 

# Explaining the white identity gender gap
Identity is a multifaceted concept. Social identity theory argues that identity salience is a function of contexts and cues. 
"White skin provides white people with the privilege to connect with other white people, to seemingly remove ourselves from an oppressed state, because a knight in shining armor protects us." [@anderson2021rebirthing, p.19].

# Theoretical Framework
As White Americans hold consequential and dominant social status, they tend to associate with their racial in-group more than any other identity. For example, White women had the opportunity to choose a white woman President during the 2016 Presidential elections. In Arlie Hochschild's words, "Trump was the identity politics candidate for white men" [@hochschild2018strangers, p. 230]. White women are also supportive of descriptive representation based on gender [@phillips2018wanting]. Among other groups such as minorities and immigrants, Trump was also openly disrespectful to women. All of this suggests that before the elections one would have expected White women to vote for Hillary Clinton. However, White women sided with Donald Trump [@junn2017trump]. 

We use the above case as a motivating example that White women's political behavior can be explained using social identity theory, according to which, individuals obtain a sense of group identity based on the social and political context. Based on that, individuals prioritize group identities that maximize their self-esteem [@tajfel1979social]. Given that women's gender makes them socially subordinate to men, white women tend to align themselves with their racial identity more than white men as this can maximize their self-esteem. This leads us to expect that White women have a higher sense of white identity than white men.        


$H_1:$ White women report a higher white identity than White men; when controlling for partisan identification.

According to social identity theory, identity formation is a function of cueing. Information about the shift in racial demography makes racial identities salient enough to trigger a perceived status threat [@craig2014more]. Extant research shows that when faced with a realistic or perceived out-group threat, highly identifying individuals of in-group feel the most threatened [@mutz2018status], which is why their tendency to protect the in-group is the strongest [@morrison2008effects]. If white women feel more associated with their racial identity, they will have a higher propensity to ......  

The political implications of racial animus are different from that of in-group solidarity [@jardina2021group]. As social identity theory postulates, identity formation is a function of cueing. Therefore, we can expect separate effects of messages evoking in-group favoritism and racial resentment. 

Ideological dispositions vary among White men and White women. White men find the messages of racial animus more appealing. White women, on the other hand, tend to be more liberal, therefore, they rely on in-group racial attitudes more than out-group racial attitudes. 

Though Republicans hold more conservative attitudes than Democrats, Democrats also hold racially prejudiced views [@krupnikov2015racial].  

$H_2:$ Hold partisanship constant, liberal women will report a higher white political identity than conservative men.
$H_3:$ Liberal women report a higher white identity because they find the message communicating white grievances more palatable than those expressing racial animus.

-->

# Study 1

```{r}
#| label: setup-block
# Title: Preliminary analyses of ANES for White identity and gender project

# Notes:
    #* Description: Rscript to perform the exploratory analyses of the 2016 and 2020 ANES for the white identity and gender project
    #* Updated: 2022-08-25
    #* Updated by: dcr 
    #* Comments: will need to clean the ideological identification variable (7-point) too. Code I have doesn't already do that for you.

# Setup
    #* Make sure you have installed the packages needed for this analysis. Uncomment and run those you don't have 
# install.packages('box')
# install.packages('dplyr')
# install.packages('modelsummary')
  #* Clean the data sets
    #** 2016  ANES
source('../code/preliminary_analyses/anes_2016_cleaning.r')
    #** 2020 ANES
source('../code/preliminary_analyses/anes_2020_cleaning.r')
  #* Modularly load the functions we'd need for analyses
box::use(
    dplyr = dplyr[select, rename],
    modelsummary = modelsummary[datasummary_skim, modelsummary],
    broom = broom[tidy],
    mice = mice[parlmice],
    Amelia = Amelia[amelia],
    ggplot2 = ggplot2[ggplot, aes, geom_point, geom_vline, geom_linerange, theme_minimal, position_dodge, labs, theme, element_text],
    #MASS = MASS[polr],
    #sjPlot = sjPlot[plot_model],
    #mediation = mediation[mediate]
    #processR = processR[regEquation],
    #lavaan = lavaan[sem]
    rstanarm = rstanarm[stan_polr, R2],
    parameters = parameters[parameters, standardize_names],
    tidybayes = tidybayes[geom_pointinterval],
    stringr = stringr[str_wrap]
)
  #* Create empty models list object
wilcox = list()
models = list()
```

For Study 1 we use the 2016 and 2020 ANES. There are two primary reasons we rely on survey data. First, we want to establish whether there is a gender gap in white identity and whether white identity is more popular among liberal women relative to conservative men - who tend to find racially resentful messages more appealing. Second, the conservative media did not exclusively focus on white identity politics but also expressed a fair amount of racial resentment as well (e.g., praising Trump's comments that conflated immigrants with gang-affiliated criminals [@derespina_2017_fox]), so we want to examine the effects of conservative media on the whole as opposed to one particular type of message.

The 2016 ANES was in the field from September, 2016 to January, 2017. As we are focusing on Whites, we subset our dataset to those who self-identify as White. This leaves us with a sample of `r nrow(anes_2016[['cleaned']])` respondents. @tbl-2016-descriptive-stats provides the unweighted characteristics of the sample. The 2020 ANES was in the field from August, 2020 until December, 2020. The sample contains `r nrow(anes_2020[['cleaned']])` White respondents. @tbl-2020-descriptive-stats provides the unweighted characteristics of the sample.

<!--
```{r}
#| label: imputation-block
#| eval: false
anes_2016[['amelia']] = amelia(anes_2016[['cleaned']], m = m, parallel = 'snow', ncpus = cores)

anes_2016[['rf_mice']] = parlmice(anes_2016[['cleaned']], m = m, method = 'rf', n.core = cores)

anes_2020[['amelia']] = amelia(anes_2020[['cleaned']], m = m, p2s = 0, parallel = 'snow', ncpus = cores)

anes_2020[['rf_mice']] = parlmice(anes_2016[['cleaned']], m = m, method = 'rf', n.core = cores)

saveRDS(anes_2016, '../data/anes_2016_imputed.RDS')
saveRDS(anes_2020, '../data/anes_2020_imputed.RDS')
```
For both samples, we examine the patterns of missingness to first establish the patterns of non-response. Figures displaying these results are contained in the appendix. We then use multiple imputation with AMELIA II, along with using random forest models to perform mulitple imputation with chained equations [see @roberts_2022_wp]. We then examine the standardized mean discrepancy between the observed values and those that are imputed. This is estimated by treating this as a covariate balancing process using entropy balancing to construct the weights for the original dataset with missigness and to contrast them with the weights for the imputed datasets [@marbach_2021_pa]. 
-->

Both surveys ask white respondents ``How important is being White to your identity". Respondents then can report the degree to which they or disagree with the statement along a 5-point Likert scale. We code responses to this question as increasing agreement. We use this as our measure of white identity as it is a self-identification with their Whiteness and, in the context of the survey, should measure this identification when primed by politics. Though it does not capture linked fate with other whites, we agree with other scholars using these datasets that it should be sufficient to measure white identity [@perez_et-al_2021_pp]. 

Both samples ask respondents about their gender identity as well. For this particular analysis we are interested in the differences between male and female identifying respondents. Though it is somewhat reductive, we code a dichotomous measure of whether the respondent is male or female. We exclude respondents who report that they do not identify with these two gender identities.

First, we start with a simple difference of means test to examine whether there are differences between men and women on white identity. Given our hypotheses, we expect that we should find that men report a higher level of white identity as this test does not control for partisanship. As white identity is coded ordinally and gender is coded dichotomously, we perform this test using a Wilcoxon non-parametric t-test as opposed to the standard t-test.

Second, we expect that once we account for partisanship, women will be more likely to report a higher white identity than men. To examine this, we fit a model using the logit$^-1$ link function. For the fitted model, we specify our prior using the $R2$ distribution as we expect that the cut points from the $k-1$ logistic regressions will explain approximately $0.3$ of the variance in the variance of our latent outcome [see @gelman_et-al_2021_cup]. We fit this model on $6$ chains with $2000$ iterations.

Third, as we expect this gender gap to be the result of the palatability of in-group messages, we expect that, when holding partisan identification constant, more liberal women will express a white identity and that more conservative men will score relatively lower on white identity as racially animus views will be more attractive.

## Results
<!--
```{r}
#| label: wilcox-model-block
#| eval: false
# Analysis
  #* diff of means
for(m in 1:m){
  wilcox[['anes_2016_amelia']][[m]] = wilcox.test(anes_2016[['amelia']][['imputatons']][[m]]$female, anes_2016[['amelia']][['imputations']][[m]]$wid, alternative = 'two.sided', conf.int = TRUE)
  #wilcox[['anes_2016_rf']][[m]] = wilcox.test(anes_2016[['rf']][[m]]$female, anes_2016[['rf']][[m]]$wid, alternative = 'two.sided', conf.int = TRUE)
  #wilcox[['anes_2020_amelia']][[m]] = wilcox.test(anes_2020[['amelia']][[m]]$female, anes_2020[['amelia']]$wid, alternative = 'two.sided', conf.int = TRUE)
  #wilcox[['anes_2020_rf']][[m]] = wilcox.test(anes_2020[['rf']][[m]]$female, anes_2020[['rf']]$wid, alternative = two.sided, conf.int = TRUE)
}

wilcox[['anes_2016_amelia_tidy']] = pool(wilcox[['anes_2016_amelia']]) |>
  tidy() |>
  mutate(year = '2016')

wilcox[['anes_2020_amelia_tidy']] = pool(wilcox[['anes_2020_amelia']]) |>
  tidy() |>
  mutate(year = '2020')

#wilcox[['anes_2016']] = wilcox.test(anes_2016[['cleaned']]$female, anes_2016[['cleaned']]$wid, alternative = 'two.sided', conf.int = TRUE) |>
#  tidy() |>
#  mutate(year = '2016')
#wilcox[['anes_2020']] = wilcox.test(anes_2020[['cleaned']]$female, anes_2016[['cleaned']]$wid, alternative = 'two.sided', conf.int = TRUE) |>
#  tidy() |>
#  mutate(year = '2020')
```
-->
```{r}
#| label: diff-of-means-block
wilcox[['anes_2016']] = wilcox.test(anes_2016[['cleaned']]$wid, anes_2016[['cleaned']]$female, conf.int = TRUE, alternative = 'two.sided')

wilcox[['anes_2020']] = wilcox.test(anes_2020[['cleaned']]$wid, anes_2020[['cleaned']]$female, conf.int = TRUE, alternative = 'two.sided')

```

The Wilcoxon test suggests that men report higher White identity than women. This occurs in 2016 (estimate = ```r sprintf('%.2f',wilcox[['anes_2016']]$estimate) ```, ```r ifelse(wilcox[['anes_2016']][['p.value']] < 0.001, print('p < 0.001'), sprintf('%.3f',wilcox[['anes_2016']]$p.value))```) and in 2020 (estimate = ```r sprintf('%.2f', wilcox[['anes_2020']]$estimate)```, ```r ifelse(wilcox[['anes_2020']][['p.value']] < 0.001, print('p < 0.001'), sprintf('%.3f',wilcox[['anes_2020']]$p.value))```). However, men tend to be more conservative than women. To test $H_1$, we examine whether White women report higher levels of White identity when we control for partisanship.

```{r}
#| label: fig-h_1
#| fig-cap: "Higher odds of women reporting white identity when controlling for Party ID"
  #* Hypothesis 1
models[['anes_2016']] = stan_polr(as.factor(wid) ~ female + pid, data = anes_2016[['cleaned']], method = 'logistic', prior = R2(0.1, 'mean'), seed = 90210, chains = 6, iter = 1000, refresh = 0)
models[['anes_2016_tidy']] = parameters(models[['anes_2016']], conf.int = 0.95, verbose = FALSE) |>
  standardize_names(style = 'broom') |>
  data.frame() |>
  mutate(model = '2016') |>
  mutate(term = case_when(term == 'female' ~ 'Female', term == 'pid' ~ 'Party ID'),
        estimate = exp(estimate),
        conf.low = exp(conf.low),
        conf.high = exp(conf.high)) |>
  filter(!is.na(term))

models[['anes_2020']] = stan_polr(as.factor(wid) ~ female + pid, data = anes_2020[['cleaned']], method = 'logistic', prior = R2(0.1, 'mean'), seed = 90210, chains = 6, iter = 1000, refresh = 0)
models[['anes_2020_tidy']] = parameters(models[['anes_2020']], conf.int = 0.95, verbose = FALSE) |>
  standardize_names(style = 'broom') |>
  data.frame() |>
  mutate(model = '2020') |>
  mutate(term = case_when(term == 'female' ~ 'Female', term == 'pid' ~ 'Party ID'),
        estimate = exp(estimate),
        conf.low = exp(conf.low),
        conf.high = exp(conf.high)) |>
  filter(!is.na(term))

models[['combined']] = rbind(models[['anes_2016_tidy']], models[['anes_2020_tidy']])

ggplot(data = models[['combined']], aes(y = term, x = estimate, xmin = conf.low, xmax = conf.high, linetype = factor(model), shape = factor(model))) +
  geom_pointinterval(position = position_dodge(width = -0.4), size = 3) +
  theme_minimal() +
  labs(linetype = 'Model', shape = 'Model', x = 'Estimate', y = '', caption = 'Data Source: 2016, 2020 ANES.\n Dots represent median draw from posterior distribution.\n Bars represent 95% credible intervals.\n Scaled as odds ratios.') +
  theme(text=element_text(size = 13, family = 'sans'))
```

@fig-h_1 visualizes the distributions of the posterior draws from the fitted model^[ @tbl-2016-2020-ordered-logits presents the full table of results.]. Both models suggest that women are more likely to report more white identity than men when we hold partisan identification constant. Specifically, for 2016 our model suggests that women are ```r sprintf('%.2f', (models[['anes_2016_tidy']]$estimate[1] - 1)* 100) ```% more likely to report a white identity when controlling for partisan identification. Our model for 2020 suggests similar patterns with women  ```r sprintf('%.2f', (models[['anes_2020_tidy']]$estimate[1] - 1)*100) ```% more likely to report a White identity when holding partisanship constant. The credible intervals of both models also suggest a low probability that, given the data, the estimated relationship in 2016 and 2020 is null.

Our analysis thus far does suggest that the gender gap exists on in-group racial attitudes and that it is dependent on whether you consider partisanship. We expect, however, that these messages are more palatable to women as a result of their tendency to be more liberal. That is, liberals are likely to reject messages expressing explicit racially resentful attitudes but are likely still attracted to racialized messages that makes them feel like they are victims too.

# Proposed Study 2

Our analyses thus far suggest that, once we control for partisanship, men and women report different levels of their attachment to their White racial identity. We argue that the mechanism explaining this phenomenon is that men and women see different appeals to their race as attractive. We propose the following experiment to offer an empirical test of our hypothesized mechanism.

We propose a three-arm online^[Which platform we use depends on our pending source of funding for the experiment, but we hope to do this either through MTurk with the CloudResearch panel or with a YouGov panel. We will limit our sample to self-identifying White respondents.] survey experiment to examine the possible differences in reported levels of attraction to in-group and out-group racial appeals. After subjects provide informed consent and answer questions about their demographics and their political views such as partisan identification, we will randomly assign participants into one of three conditions: (1) a placebo condition, (2) an in-group racial appeal condition, and (3) an out-group racial appeal condition. In the three conditions, the participants will be presented with a text-based vignette that discuss a recent speech delivered by a candidate in a Judicial election^[**Bolded** sections within brackets represent the parts of the vignette that vary from condition. [**Placebo Condition**, **In-Group Condition**, **Out-group Condition**]]:

> In a recent speech, [NAME REDACTED BY RESEARCHERS], a candidate for an upcoming Judicial election, got onto the issue of community. They said "You've got these **[people who have little concern for fostering a community with their neighbors/Mexicans who move in, take jobs, and change the community/Illegal aliens who don't care about anyone about themselves]**". They go on to say, "We need communities **[that foster neighborly mutual trust and respect/that uphold the values that have traditionally preserved the values of this country/that keep those who are unlike us out]**." 

Our rationale for our vignette choices is as follows. We first chose for this to be a candidate in a local, judicial election. The reason we do this is out of concern for balancing external and internal validity. As judicial elections tend to be low-information and non-partisan affairs, we did this so that the vignette appeared more natural than one for a high-information and partisan election where subjects would expect information about the candidate's partisanship, thus potentially acting as a confound, or requiring that we choose a real political candidate. We also chose to focus the statements by the fictional candidate on discussing issues surrounding "community". This avoids the problem of subjects detecting partisan policy stances from the message, beyond those that are implied by the racial attitudes expressed. We focus the in-group sentiments to be about changes to the status quo which often goes along with feelings that Whites are now victims of a rapidly changing world and a racial hierarchy. In turn, we focus the out-group sentiments to be much more focused on dehumanizing those from the racial out-group and to keep them segregated.

After receiving the manipulation, we intend to present the participants with an open-ended manipulation check:
- The candidate that you just read about was running for office in which state? 

To further detect insincere responses, we follow the advice of @kennedy_et-al_2021_poq and intend to ask the subjects four open-ended responses about their political views on various issues throughout the administration of the experiment. Using the responses from these questions, we can examine those who not only provide duplicate but also insincere responses and exclude them from our analysis. We intend to do this in conjunction with the removal of those who complete the study in a higher-and-lower-than average amount of time and those who take the survey from a duplicated IP address.

After receiving the manipulation and manipulation check, we intend to ask participants the following questions:

1. After reading this news coverage of this candidate, would you be inclined to support them? (options: Support them, Would not support them, do not feel more or less inclined to support them)
2. Please indicate the extent to which you agree or disagree with the views expressed by the candidate. (options: Strongly agree, agree, slightly agree, neither agree nor disagree, slightly disagree, disagree, strongly disagree)
3. Please indicate the extent to which you agree or disagree with the view that this community is a victim to current trends. (options: Strongly agree, agree, slightly agree, neither agree nor disagree, slightly disagree, disagree, strongly disagree)
4. WHITE IDENTITY SCALE
5. RACIAL RESENTMENT SCALE

The goal of question 1 is to examine whether the message is attractive enough to shift the subject's desire to "publicly" express support of the candidate and their views. Question 2 is designed to examine how attractive the message itself is. Question 3 is designed to dig into their view that this message communicates a victimization of the community by another racial group. We will then present respondents with Jardina's [-@jardina_2019_cup] white identity scale and with the racial resentment scale. 


# Discussion

In this manuscript we develop a theoretical framework that explain a phenomenon that we refer to as the racialized gender gap. We argue that the racialized gender gap is the result of a difference in demand for white women to balance a number of considerations when evaluating the appeal of different racialized political messages. We argue that the result of this balancing act leads white women to be more likely to find in-group oriented attitudes and messages of white victimhood appealing relative to men who do not have to balance the same considerations and are more likely to find out-group messages of racial animus more appealing. 

While in this current draft, we do not test this mechanism. We do use the 2016 and 2020 ANES to establish that the racialized gender gap does indeed exist. We also propose an experimental design that we hope will be a good test of our mechanism. We appreciate feedback to strengthen our project. 

<!--
Academics and political observers are documenting a shift in the approach that Republicans and conservatives are taking in their messages about race. Coverage among, large, nationwide news organizations connect the messaging strategy among Republicans to play up white victimhood as a way to justify their policy positions [@samuels_lewis_2022_fte]. Academics make similar claims connecting white identity with support for Donald J. Trump in the 2020 Presidential election [@jardina_et-al_2020]. Some non-academic analyses connect that Trump frames individual problems as resulting from widespread discrimimination against Whites [@harwood_2022_cnn]. These elite messages stand in stark contrast to those that express racially resentful views of non-Whites. As these racially resentful messages were popular, in particular, during Barack Obama's administration among conservatives [@tesler_2016], it appears that there has been a recent shift in Republican messaging. This is not to say, however, that Trump and other politicians do not express racially resentful views; he does. On balance, however, White grievance politics appears to play a larger role in elite messages than it used to. 

As partisan identification represents a "mega-identity" that encapsulates many other social identities [@mason_2015_ajps], evidence suggests that attitudes toward racial groups and partisans are strongly influential on each other [@westwood_peterson_2022_pb]. These messages play well among Whites - particularly among conservative men. Though some attribute the shift in racial attitudes among the American public to Democrats who hold racially conservative views [@sides_et-al_2018], it remains unclear as to why Republicans believe a shift in messages that focus on the plight of one racial group as beneficial. 

We argue that the shift from messages which focus on individual attributions of blame for poor circumstances among non-Whites and hostility toward systematic-level efforts to address it [see @winter_2008] to those that express concern with system-wide bias against Whites, may be partially attributable to the Republican party's concern with attracting White female voters. In the 1972 Presidential election, 38% of women voted for the Democratic candidate, but in 2012 52% of women did [@chaturvedi_2016_pew]. However, about 52% of White women voted for Trump in the 2016 Presidential election [@junn_2017_pgi]. Some suggest that conservative men becoming much more conservative is to blame for the opening of the gender gap [@kaufmann_petrocik_1999_ajps]. Messages of racial resentment were more popular among racially conservative men CITATION. Though we cannot get into the heads of political strategists for the party, it is reasonable to assume that the party will try to expand its voting base and that it will focus on messages that are popular among segments of the population they want to attract. This means we will examine whether it is the case that these in-group oriented racialized messages are popular among White women.
-->

{{< pagebreak >}}

# Appendix

## Descriptive statistics

{{< pagebreak >}}

```{r}
#| label: tbl-2016-descriptive-stats
#| tbl-cap: 2016 Descriptive statistics
# Tables
    #* Descriptive statistics
        #** 2016 ANES
anes_2016[['cleaned']] |>
    select(wid, female, pid, conserv_media, gender_discrim) |>
    rename(`White identity` = wid,
            `Female` = female,
            `Party ID` = pid,
            `Conservative media` = conserv_media,
            `Gender discrimination` = gender_discrim) |>
    datasummary_skim(histogram = FALSE, notes = c('Note: Unique column includes NA values.', 'Data source: 2016 American National Election Study.'), output = 'latex')
```

```{r}
#| label: tbl-2020-descriptive-stats
#| tbl-cap: 2020 Descriptive statistics
        #** 2020 ANES
anes_2020[['cleaned']] |>
    select(wid, female, pid, conserv_media, gender_discrim) |>
    rename(`White identity` = wid,
            `Female` = female,
            `Party ID` = pid,
            `Conservative media` = conserv_media,
            `Gender discrimination` = gender_discrim) |>
    datasummary_skim(histogram = FALSE, notes = c( 'Note: Unique column includes NA values.', 'Data source: 2020 American National Election Study.'), output = 'latex')
```

<!--
## Patterns of missingness
```{r}
#| label: fig-missing-pattern_2016
#| fig-cap: Patterns of missing values - 2016 ANES
#| eval: false
#| echo: false
#| warning: false
#| message: false
explanatory_2016 = colnames(anes_2016[['cleaned']])[colnames(anes_2016[['cleaned']]), != `White identity`]
anes_2016[['cleaned']] |>
  mutate(
    `White identity` = factor(wid, labels = c('Not at all important', 'A little important', 'Moderately important', 'Very Important', 'Extremely important')),
    `Female` = factor(female, labels = c('Male', 'Female')),
    `Party ID` = factor(pid, labels = c('Strong Democrat', 'Democrat', 'Lean Democrat', 'Independent', 'Lean Republican', 'Republican', 'Strong Republican'))
  ) |>
  missing_pairs(dependent = 'White identity', explanatory = explanatory_2016) + labs(caption = 'Data source: 2016 American National Election Study.')

```

```{r}
#| label: fig-missing-pattern-2020
#| fig-cap: Patterns of missing values - 2020 ANES
#| eval: false
explanatory_2020 = colnames(anes_2020[['cleaned']])[colnames(anes_2020[['cleaned']]), != `White identity`]

anes_2020[['cleaned']] |>
  mutate(
    `White identity` = factor(wid, labels = c('Not at all important', 'A little important', 'Moderately important', 'Very important', 'Extremely important')),
    `Female` = factor(female, labels = c('Male', 'Female')),
    `Party ID` = factor(pid, labels = c('Strong Democrat', 'Democrat', 'Lean Democrat', 'Independent', 'Lean Republican', 'Republican', 'Strong Republican'))
  ) |>
  missing_pairs(dependent = 'White identity', explanatory = explanatory_2020) + labs(caption = 'Data source: 2020 American National Election Study.')
```
-->


## Models


```{r}
#| label: tbl-2016-2020-ordered-logits
#| tbl-cap: Women have higher log-likelihood of reporting high white identity when controlling for Party ID
gm = list(list('raw' = 'nobs', 'clean' = 'N', fmt = 0))
model_table = list(
  '2016' = models[['anes_2016']],
  '2020' = models[['anes_2020']]
)
modelsummary(model_table, output = 'latex', gof_map = gm, coef_map = c('female' = 'Female', 'pid' = 'Party ID', '1|2' = 'Threshold 1', '2|3' = 'Threshold 2', '3|4' = 'Threshold 3', '4|5' = 'Threshold 4'), statistic = 'conf.int', notes = list('Brackets represent credible intervals at the 95-percent level.','Point estimates are the logged odds ratios.', 'Data source: 2016, 2020 American National Election Study.'))
```

{{< pagebreak >}}